Diversity and evolution of centromere repeats in the maize genome

### 2014-04-19 centc.tex
minor edits

\subsection*{CentC homoplasious mutations simulation}

@@

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
%  \includegraphics[width=0.75\textwidth]{circos.png}
  \includegraphics{circos.png}
\caption{CentC repeat location in relation to the maize subgenomes.  High confidence regions are colored with darker colors while low confidence regions are colored with lighter colors.  Breakpoints between the subgenomes remain uncolored to indicate uncertainty.}
\label{circos}    
\end{figure}

@@

The decreased genetic distance among CentCs in local clusters on chromosome 2 and 5 suggest that many of the genetic groupings discovered in our genome-wide analysis should correspond to local clusters of repeats. We see no evidence of this, however, as repeats within individual clusters are frequently found in different genetic groups as defined by PCoA (Figure \ref{pcoa}).  A comparison of all pairs of CentC reveals a likely explanation: of the $\approx 74$ million possible pairs,  approximately 6 million share $\geq 2$ mutations different from the genome-wide consensus, likely grouping in genetic clusters despite their physical distance.  Looking at several triplets at random from our alignment confirms that two sequences in one PCoA assignment share greater pairwise identity than two sequences adjacent to one another in different PCoA groups.  A simple forward simulation (see \ref{methods} \pb{guessing you want link to github here, or link to section in methods?}) suggests this pattern could be due entirely to homoplasy rather than long-distance movement of CentC repeats.  By stochastically applying mutations to an initially homogeneous group of repeat sequences, we find that plausible parameter values produce $\approx 10$ million pairs of repeats sharing $\geq 2$ mutations.

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{TW_CentCgraph.jpeg}
\caption{Presence of CentC in each of the heirarchical groups.  The 58 clusters found to be statistically significant in forming genetic groups are represented on the x-axis and chromosome of origin for CentC's on the y-axis. The intersection shows the number of CentC's in the group found on that chromosome, scaled logarithmically.  CentC counts from chromosomes whose centromeres were derived from subgenome 1 are colored green and those from subgenome 2 are colored brown.
}
\label{pcoa}    
\end{figure}

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{Figure3_CentC_Abundance.png}
\caption{(a) Mb of CentC in genomic libraries of maize and teosinte.  Box plots show data from \citet{Chia2012}.  Points show data for maize inbred B73 and the teosinte \emph{Z. luxurians} from \citet{Tenaillon2011}.  For comparison, the data point of maize inbred B73 in \citet{Chia2012} is shown with a tick mark on the box plot. (b) FISH images from an F1 hybrids between maize and the teosinte \emph{parviglumis}, and maize and the teosinte \emph{Z. luxurians}.  Upper row shows the karyotype for the teosinte and the lower row shows the maize karyotype.  Red=CentC, Green=CRM. 
}
\label{abundance}    
\end{figure}

@@

Shotgun sequence data from the maize HapMap v2 \citep{Chia2012}, reveals a significantly greater abundance of CentC in teosinte than in domesticated maize (p$<0.01$; Figure \ref{abundance}). Teosintes have more CentC than inbred maize.  Further support for differences between maize and its wild relatives comes from shotgun sequence data from \emph{Z. luxurians} \citep{Tenaillon2011}.  Analysis of these data find nearly twice as much CentC in \emph{Z. luxurians} as the maize inbred B73.  To corroborate these results, we performed fluorescent in-situ hybridization of F1 crosses between inbred maize and teosinte to determine if cytological observations agreed with our sequencing findings.  Cytology needed to be performed on an F1 cross so that we could compare relative probe fluorescence of the chromosomes within a single individual.  FISH data supports our observation that the teosintes \emph{parviglumis} and \emph{Z. luxurians} have more CentC than inbred maize (Figure \ref{abundance}). \jri{add in this figure, and better quality images of first two} \pb{of the FISH?  thats the best quality i have insert slantyface}.

@@

Our study traces the changes in centromere repeat genetic relatedness across maize chromosomes to show how ancient tetraploidy and subsequent evolution has impacted centromeres.  Most interestingly, we show that most large arrays of CentC in the maize reference genome derive from maize1 subgenome, which is known to have lower gene loss and higher average expression of its genes than maize subgenome 2.  Regulatory microRNA’s are known to correspond to centromeric repeats \citep{ReinhartBartel2002}.  Therefore, it may be worthwhile to further investigate the potential correlation between centromere retention and gene expression as the phenomenon of higher expression of genes from one ancestor may be very common across plants (\emph{A. thaliana} \cite{Cheng2012} and \emph{Gossypium} Kapp and Wendel unpublished).  \pb{Two comments about this: first, its a terrible sentence so help make this better, because i think the point is really nice.  second, the best citations i found for cotton was an unpublished ref that is still not published from what i can gather.  advice? e.g. \url{http://www.plosone.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pone.0036442} also check out papers that cite Schabs: \url{http://scholar.google.com/scholar?cites=5313714612108479532&as_sdt=2005&sciodt=0,5&hl=en} } There are several possible explanations for the differential contribution of subgenome 1 to the diversity of CentC in the extant maize  genome. We put forth several hypotheses that may explain the strong bias for CentC to be located within a block from subgenome 1. Subgenome 2 may have had severely reduced quantities of CentC and therefore never contributed equally.  The CentC from subgenome 2 may have decayed rapidly and thus less of it remains.  Lastly, the CentC from subgenome 1 may have expanded in copy number since the allopolyploidy event.  Though we are unable to conclusively identify the reason, the lack of a large number of identical tandem duplicates  and lack of close proximity highest genetic relatedness pairs suggests that any large scale expansion of CentC has not been recent.  Also, CentC’s in subgenome 2 cluster do not appear to have accumulated excess mutation, suggesting that they are not decaying more rapidly than their subgenome 1 counterparts.  We therefore believe that it is most likely that the subgenomes had an unequal contribution of CentC at the formation of the allopolyploid, though more research about the ancient parents is required.

We also explored how genetic relatedness between CentC’s correlated with location along the physical map.  According to our PCoA analyses, CentC copies across the genome do not form distinct genetic groups correlating with past origin (Figure \ref{pcoa}).  The lack of subgenome or chromosome grouping in modern CentC suggests a little differentiation between the repeats since the ancestral divergence.  \jri{the fact that repeats in Trip are the same suggests no differentiation in ancestors right? if they had homogenized (read: changed) since ancestors, why would trip == maize?} \pb{unless trip and maize both went with the CentC most like one of the 2 ancestors, we can't know without knowing the ancestral repeat.  I mean I would venture to say that the repeat slated to "win" the fight in maize is also the one I would place my bets on in trip} Given the lack of a genome wide pattern, we also wanted to investigate genetic relatedness on individual chromosomes.  We hypothesized that most copies of CentC arose from local duplication rather than transposition, and therefore the genetic distance between CentC’s would be lowest across CentC’s within a large tandem array.  \pb{This part is ugly, and needs help} From the reference genome, we chose to investigate chromosomes 2 and 5, since they have been sequenced from end to end, allowing for inferences about the relationship between diversity and physical proximity \citep{Wolfgruber2009}.  Furthermore, the role of the repeat arrays on chromosomes 2 and 5 appear very different, as the largest array on chromosome 2 interacts with the kinetochore, while no array on chromosome 5 does \citep{Wolfgruber2009}.  We find that repeats on chromosomes 2 and 5 are most highly genetically related to neighboring copies (Figure \ref{heatmap}).  This relationship was recapitulated in analyses using SpaGeDi, where CentC’s within approximately 10-50KB of one another were more genetically similar (Supplemental Data \ref{Supp_SpagediChr02} and \ref{Supp_SpagediChr05}).  Though we did not investigate the relationship between CentC location and genetic similarity on the other chromosomes due to incomplete sequencing of centromeres, we observe that many CentC’s within an array fall into the same  significant grouping in our Tracy-Widom clustering analysis, suggesting that high local similarity of CentC’s is a genome-wide phenomenon.  This local relatedness suggests that most CentC is evolving through tandem duplication and not a long distance mechanism such as transposition that had been previously suggested \citep{Shi2010}.  Concerning chromosomes 2 and 5, the higher local relatedness of all CentC clusters regardless of presence within the kinetochore suggests that CentC interaction with kinetochore proteins does not have a detectable change on their rates of evolution.

@@

When studying CentC changes through domestication, our findings show that the repeat has experienced a decrease in copy number over time without a noticeable change in the structure of the repeat arrays. Using PacBio long read sequencing, we confirm genome-wide the observation we see on chromosomes 2 and 5, that most copies of CentC exist in large tandem arrays in all \emph{Zea} taxa.  \jri{i think the discussion should largely mirror the results. this doesn’t go in the domestication part really.}  \pb{the pacbio is no where in the results, just methods.  Ok here in discussion?}From short read sequencing data, we show that modern maize has reduced genomic abundance of CentC when compared to the teosintes, a finding confirmed through FISH (Figure \ref{abundance}).  Both teosintes within the species \emph{mays} and its sister species \emph{luxurians} have elevated levels of CentC.  Lower levels of centromere repeats in inbred maize contrasts previous studies that characterized the changing abundance of other common repeats.  For example, the abundance of most transposable element families increased after domestication \citep{Chia2012}.  Knowing that TE abundance increases in domesticated maize, we might have expected CentC content to increase as well if centromere size had to expand alongside most other repetitive content.  Alternatively, due to their structural role in kinetichore formation, we might have also expected centromere repeats to be largely excluded from the genome wide fluctuations in repetitive content assuming that selection exists to maintain centromere size.  Instead, the decrease in CentC content may indicate an active process of removing CentC.  We hypothesize that the removal of CentC content may correlate with genome size, as the sharp decrease in knob content through domestication actually led to an overall smaller genome size in inbred maize. The correlation between  functional  centromere size and genome size has been observed in grass species \citep{Zhang2012}. \jri{the problem here is kelly shows functional cent. size, not repeat array size.   } \pb{i dont think this changes the point, CentC, whether in or out the kinetichore, seems to maintain the same relatedness/diversity, not sure this is addressed so left comments in here} Further investigation would be required to discover whether only those copies of CentC outside of the active kinetochore are being deleted.

\pb{Michelle is pretty adamant that this isnt a TE. As i found, no GAG or POL, but still a repeat at ends.  Strike the whole thing, or worthwhile mentioning something about it?} We also identify the cent4 repeat as the LTR retrotransposon RLX\_sela.  Characterization of the TE consensus sequences shows that it contains a 232bp LTR on both ends but lacks any of the protein sequences for autonomous transposition, such as GAG and POL complexes.  Furthermore, RLX\_sela has several \jri{why brackets? is this uncerstain?}\pb{had brackets because its unclear how I want to group these and set lengths of them, so i settle for several.} regions showing $>60$bp of homology to the 180bp knob repeat. The lack of binding between CenH3 and RLX\_sela seen in the ChIP data further suggests that this repeat is likely pericentromeric rather and not involved in centromere function, a result echoed by  \citet{Jin2004}, who showed that Cent4 probes lagging behind CentC probes in cell division. Previous work in rice has documented the presence of nonautonomous LTR retrotransposons  in or near the centromere \citep{Jiang2002}.  Given that the centromere on chromosome 4 shows signs of strong selection during domestication \citep{Hufford2012} \jri{doublecheck i'm not misremembering this please} \pb{Super high peak in landracevsimproved just below centromere, and reasonable peak around centromere in teovslandrace. However, I am pretty sure there are enough peaks to say sure, one of those is this}, it is plausible that the increase in abundance of Cent4 seen in domesticated maize is due to the effects of linked selection on a locus important for domestication rather than novel insertions or duplications.

@@

\begin{acknowledgements}
We wish to thank Pacific Biosciences for sequencing resources.  We thank Lauren Sagara, Gernot Presting,  NSF summer exchange program interns Cesar Alvarez Mejia, Aurelio Hernandez Bautista, and Siddharth Bhadra-Lobo for helpful discussion.

>>>>

\subsection*{Simulation of homoplasious mutations}

@@

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
%  \includegraphics[width=0.75\textwidth]{circos.png}
  \includegraphics{circos.png}
\caption{CentC repeat location in relation to the maize subgenomes.  High confidence regions are colored with darker colors while low confidence regions are colored with lighter colors.  Breakpoints between the subgenomes remain uncolored to indicate uncertainty. \jri{fix legend. what is middle outer inner ring?}}
\label{circos}    
\end{figure}

@@

The decreased genetic distance among CentCs in local clusters on chromosome 2 and 5 suggest that many of the genetic groupings discovered in our genome-wide analysis should correspond to local clusters of repeats. We see no evidence of this, however, as repeats within individual clusters are frequently found in different genetic groups as defined by PCoA (Figure \ref{pcoa}).  A comparison of all pairs of CentC reveals a likely explanation: of the $\approx 74$ million possible pairs,  approximately 6 million share $\geq 2$ mutations different from the genome-wide consensus, likely grouping in genetic clusters despite their physical distance.  Looking at several triplets at random from our alignment confirms that two sequences in one PCoA assignment share greater pairwise identity than two sequences adjacent to one another in different PCoA groups.  A simple forward simulation (see Methods) suggests this pattern could be due entirely to homoplasy rather than long-distance movement of CentC repeats.  By stochastically applying mutations to an initially homogeneous group of repeat sequences, we find that plausible parameter values produce $\approx 10$ million pairs of repeats sharing $\geq 2$ mutations.

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{TW_CentCgraph.jpeg}
\caption{Presence of CentC in each of the heirarchical groups.  The 58 clusters found to be statistically significant in forming genetic groups are represented on the x-axis and chromosome of origin for CentC's on the y-axis. The intersection shows the number of CentC's in the group found on that chromosome, scaled logarithmically.  CentC counts from chromosomes whose centromeres were derived from subgenome 1 are colored green and those from subgenome 2 are colored brown. \jri{low quality fig. use pdf not png or jpeg.}
}
\label{pcoa}    
\end{figure}

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{Figure3_CentC_Abundance.png}
\caption{(a) Mb of CentC in genomic libraries of maize and teosinte.  Box plots show data from \citet{Chia2012}.  Points show data for maize inbred B73 and the teosinte \emph{Z. luxurians} from \citet{Tenaillon2011}.  For comparison, the data point of maize inbred B73 in \citet{Chia2012} is shown with a tick mark on the box plot. (b) FISH images from an F1 hybrids between maize and the teosinte \emph{parviglumis}, and maize and the teosinte \emph{Z. luxurians}.  Upper row shows the karyotype for the teosinte and the lower row shows the maize karyotype.  Red=CentC, Green=CRM. \jri{center these figs}
}
\label{abundance}    
\end{figure}

@@

Shotgun sequence data from the maize HapMap v2 \citep{Chia2012}, reveals a significantly greater abundance of CentC in teosinte than in domesticated maize (p$<0.01$; Figure \ref{abundance}). Teosintes have more CentC than inbred maize.  Further support for differences between maize and its wild relatives comes from shotgun sequence data from \emph{Z. luxurians} \citep{Tenaillon2011}.  Analysis of these data find nearly twice as much CentC in \emph{Z. luxurians} as the maize inbred B73.  To corroborate these results, we performed fluorescent in-situ hybridization of F1 crosses between inbred maize and teosinte to determine if cytological observations agreed with our sequencing findings.  Cytology needed to be performed on an F1 cross so that we could compare relative probe fluorescence of the chromosomes within a single individual.  FISH data supports our observation that the teosintes \emph{parviglumis} and \emph{Z. luxurians} have more CentC than inbred maize (Figure \ref{abundance}).

@@

Our study traces the changes in centromere repeat genetic relatedness across maize chromosomes to show how ancient tetraploidy and subsequent evolution has impacted centromeres.  Most interestingly, we show that most large arrays of CentC in the maize reference genome derive from maize1 subgenome, which is known to have lower gene loss and higher average expression of its genes than maize subgenome 2.  \jri{cite needed} Regulatory microRNA’s are known to correspond to centromeric repeats \citep{ReinhartBartel2002}.  Therefore, it may be worthwhile to further investigate the potential correlation between centromere retention and gene expression as the phenomenon of higher expression of genes from one ancestor may be very common across plants (\emph{A. thaliana} \cite{Cheng2012} and \emph{Gossypium} Kapp and Wendel unpublished).  \pb{Two comments about this: first, its a terrible sentence so help make this better, because i think the point is really nice.  second, the best citations i found for cotton was an unpublished ref that is still not published from what i can gather.  advice? e.g. \url{http://www.plosone.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pone.0036442} also check out papers that cite Schabs: \url{http://scholar.google.com/scholar?cites=5313714612108479532&as_sdt=2005&sciodt=0,5&hl=en} } There are several possible explanations for the differential contribution of subgenome 1 to the diversity of CentC in the extant maize  genome. We put forth several hypotheses that may explain the strong bias for CentC to be located within a block from subgenome 1. Subgenome 2 may have had severely reduced quantities of CentC and therefore never contributed equally.  The CentC from subgenome 2 may have decayed rapidly and thus less of it remains.  Lastly, the CentC from subgenome 1 may have expanded in copy number since the allopolyploidy event.  Though we are unable to conclusively identify the reason, the lack of a large number of identical tandem duplicates  and lack of close proximity highest genetic relatedness pairs suggests that any large scale expansion of CentC has not been recent.  Also, CentC’s in subgenome 2 cluster do not appear to have accumulated excess mutation, suggesting that they are not decaying more rapidly than their subgenome 1 counterparts.  We therefore believe that it is most likely that the subgenomes had an unequal contribution of CentC at the formation of the allopolyploid, though more research about the ancient parents is required.

We also explored how genetic relatedness between CentC’s correlated with location along the physical map.  According to our PCoA analyses, CentC copies across the genome do not form distinct genetic groups correlating with past origin (Figure \ref{pcoa}).  The lack of subgenome or chromosome grouping in modern CentC suggests a little differentiation between the repeats since the ancestral divergence.  Given the lack of a genome wide pattern, we also wanted to investigate genetic relatedness on individual chromosomes.  We hypothesized that most copies of CentC arose from local duplication rather than transposition, and therefore the genetic distance between CentC’s would be lowest across CentC’s within a large tandem array.  \pb{This part is ugly, and needs help} From the reference genome, we chose to investigate chromosomes 2 and 5, since they have been sequenced from end to end, allowing for inferences about the relationship between diversity and physical proximity \citep{Wolfgruber2009}.  Furthermore, the role of the repeat arrays on chromosomes 2 and 5 appear very different, as the largest array on chromosome 2 interacts with the kinetochore, while no array on chromosome 5 does \citep{Wolfgruber2009}.  We find that repeats on chromosomes 2 and 5 are most highly genetically related to neighboring copies (Figure \ref{heatmap}).  This relationship was recapitulated in analyses using SpaGeDi, where CentC’s within approximately 10-50KB of one another were more genetically similar (Supplemental Data \ref{Supp_SpagediChr02} and \ref{Supp_SpagediChr05}).  Though we did not investigate the relationship between CentC location and genetic similarity on the other chromosomes due to incomplete sequencing of centromeres, we observe that many CentC’s within an array fall into the same  significant grouping in our Tracy-Widom clustering analysis, suggesting that high local similarity of CentC’s is a genome-wide phenomenon.  This local relatedness suggests that most CentC is evolving through tandem duplication and not a long distance mechanism such as transposition that had been previously suggested \citep{Shi2010}.  Concerning chromosomes 2 and 5, the higher local relatedness of all CentC clusters regardless of presence within the kinetochore suggests that CentC interaction with kinetochore proteins does not have a detectable change on their rates of evolution.

@@

When studying CentC changes through domestication, our findings show that the repeat has experienced a decrease in copy number over time without a noticeable change in the structure of the repeat arrays. Using PacBio long read sequencing, we confirm genome-wide the observation we see on chromosomes 2 and 5, that most copies of CentC exist in large tandem arrays in all \emph{Zea} taxa.  \jri{i think the discussion should largely mirror the results. this doesn’t go in the domestication part really.}  \pb{the pacbio is no where in the results, just methods.  Ok here in discussion?} \jri{nope, need pacbio in results, even if briefly} From short read sequencing data, we show that modern maize has reduced genomic abundance of CentC when compared to the teosintes, a finding confirmed through FISH (Figure \ref{abundance}).  Both teosintes within the species \emph{mays} and its sister species \emph{luxurians} have elevated levels of CentC.  Lower levels of centromere repeats in inbred maize contrasts previous studies that characterized the changing abundance of other common repeats.  For example, the abundance of most transposable element families increased after domestication \citep{Chia2012}.  Knowing that TE abundance increases in domesticated maize, we might have expected CentC content to increase as well if centromere size had to expand alongside most other repetitive content.  Alternatively, due to their structural role in kinetichore formation, we might have also expected centromere repeats to be largely excluded from the genome wide fluctuations in repetitive content assuming that selection exists to maintain centromere size.  Instead, the decrease in CentC content may indicate an active process of removing CentC.  We hypothesize that the removal of CentC content may correlate with genome size, as the sharp decrease in knob content through domestication actually led to an overall smaller genome size in inbred maize. The correlation between  functional  centromere size and genome size has been observed in grass species \citep{Zhang2012}. \jri{the problem here is kelly shows functional cent. size, not repeat array size.   } \pb{i dont think this changes the point, CentC, whether in or out the kinetichore, seems to maintain the same relatedness/diversity, not sure this is addressed so left comments in here} Further investigation would be required to discover whether only those copies of CentC outside of the active kinetochore are being deleted.

\pb{Michelle is pretty adamant that this isnt a TE. As i found, no GAG or POL, but still a repeat at ends.  Strike the whole thing, or worthwhile mentioning something about it? } \jri{i say we keep in, but with a more nuanced explanation.  basically we can say it is tandem. i say some sleuthing may be in order -- simple blast hits find it in oryza and vitis telmore repeats...}We also identify the cent4 repeat as the LTR retrotransposon RLX\_sela.  Characterization of the TE consensus sequences shows that it contains a 232bp LTR on both ends but lacks any of the protein sequences for autonomous transposition, such as GAG and POL complexes.  Furthermore, RLX\_sela has several \jri{why brackets? is this uncerstain?}\pb{had brackets because its unclear how I want to group these and set lengths of them, so i settle for several.} regions showing $>60$bp of homology to the 180bp knob repeat. The lack of binding between CenH3 and RLX\_sela seen in the ChIP data further suggests that this repeat is likely pericentromeric rather and not involved in centromere function, a result echoed by  \citet{Jin2004}, who showed that Cent4 probes lagging behind CentC probes in cell division. Previous work in rice has documented the presence of nonautonomous LTR retrotransposons  in or near the centromere \citep{Jiang2002}.  Given that the centromere on chromosome 4 shows signs of strong selection during domestication \citep{Hufford2012} \jri{doublecheck i'm not misremembering this please} \pb{Super high peak in landracevsimproved just below centromere, and reasonable peak around centromere in teovslandrace. However, I am pretty sure there are enough peaks to say sure, one of those is this}, it is plausible that the increase in abundance of Cent4 seen in domesticated maize is due to the effects of linked selection on a locus important for domestication rather than novel insertions or duplications.

@@

\begin{acknowledgements}
\jri{add vince, anne lorant?}
We wish to thank Pacific Biosciences for sequencing resources.  We thank Lauren Sagara, Gernot Presting,  NSF summer exchange program interns Cesar Alvarez Mejia, Aurelio Hernandez Bautista, and Siddharth Bhadra-Lobo for helpful discussion.
